Predicting the effects of character displacement on food-web dynamics

### 2019-03-18 manuscript_ECD_model.Rmd
added edits to main text and supplementary information.

Previous studies have analyzed the evolution of consumer attack rates in the first two foraging scenarios using an Adaptive Dynamics approach, with the general result being divergent character displacement [@Lawlor1976; @Abrams1986]. 
I say consumers have undergone divergent character displacement if there evolved attack rates are more specialized, measured as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer.
I also used an Adaptive Dynamics approach to analyze character displacement in the third foraging scenario, and I too observed divergent character displacement (detailed analysis in Appendix S1 of Supplementary Information).

@@

Given that divergent character displacement was an expected result, I focus here on its consequences for food-web dynamics across these foraging scenarios.
To do this, I analyzed differences in resource abundances and food-web stability at equilibrium. 
An equilibrium is reached when the four rates of change in the above equations are 0, and solving the system at this point gives equilibrium abundances for each resource ($\hat R_1,\hat R_2$) and each consumer ($\hat C_1,\hat C_2$). 
At the coexistence equilibrium (i.e., where $\hat R_1,\hat R_2, \hat C_1,\hat C_2 > 0$), I calculated resource abundances and performed a local stability analysis [@Otto2007]. 
This stability analysis derives the dominant eigenvalue, $\lambda$, which determines whether (and how readily) the food web will return to equilibrium after a small perturbation.

When possible, I derived analytical expressions for the relationship between consumer attack rates and food-web dynamics. 
To do this, I simplified the model by assuming that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $h=h_{11}=h_{12}=h_{21}=h_{22}$; $m=m_1=m_2$), except that consumer attack rates and their habitat preferences (if present) are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$; $w_{ii}=w_{11}=w_{22}$). 
Controlling for other sources of variability allowed me to isolate the general effects of divergent character displacement.
All mathematical derivations were conducted in @Mathematica.

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w_{ii}>0.5$ and $a_{ii}>a_{ij}$), then the effective attack rate of consumers ($wa_{ii}+(1-w_{ii})a_{ij}$) will always increase, regardless of the tradeoff (Fig. \ref{fig:plot_fig3}). 
Thus, divergent character displacement always results in resource suppression (Fig. \ref{fig:plot_fig3}). 
Note that the shape of the tradeoff can modify the effect of character displacement. 
This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached an evolutionary stable strategy (ESS, triangles in Fig. \ref{fig:plot_fig3}). 
In contrast, resource abundances reach a similar value when consumers evolve in the presence of a competitor (circles in Fig. \ref{fig:plot_fig3}), because character displacement tends to reach a constraint of complete specialization. 
It is worth noting that resource abundances are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in Fig. \ref{fig:plot_fig3}). 
This is because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in Fig. \ref{fig:plot_fig3}).

As seen previously, the effect of character displacement on resource abundances qualitatively corresponds to its effect on food-web stability (Fig. \ref{fig:plot_fig3}C). 
Specifically, divergent character displacement decreases food-web stability, regardless of the tradeoff in attack rates.
This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (Fig. \ref{fig:plot_fig3}C). 
For example, when the tradeoff is concave up (orange), then the four-species (small circle) food web actually starts off as more stable than the three-species one (small triangle); however, this pattern switches by the end of the eco-evolutionary simulation (large points).

@@

In the third foraging scenario, I observed the same general effect of divergent character displacement as the previous scenario (resources in different habitats, but linear functional response).
<!--Divergent character displacement in this foraging scenario has the same general effect on food-web dynamics as the previous scenario (resources in different habitats, but linear functional response). 
<!--In terms of food-web dynamics, I observe the same general effect of ecological character displacement as before when resources occurred in distinct habitats, but consumers had a simpler functional response. -->
This is because resource abundances at equilibrium are governed by a similar dynamic (derived in Appendix S**?**):
<!--
Even after adding more biological realism, I still observe the same qualitative relationship between character displacement and resource suppression (Fig. \ref{fig:McCann_Plot} in Appendix). This is because resource competition always leads to character divergence and equilibrium resource abundances are governed by similar dynamics  (derivation in supplementary material):
-->

@@

```{r Stability_McCann, include=FALSE, fig.cap="\\label{fig:Stability_McCann}Effect of character displacement on local stability when consumers exhibit a more realistic functional response."}
Stability_McCann <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen, color=factor(n))) + 
  geom_line() +
  geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend = FALSE) +
  geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(bquote(atop("Food-web stability","("*-lambda[max]*")"))) +
  #ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r McCann Model, include=FALSE}
plot_McCann_model <- plot_grid(
  McCann_attack    + theme(legend.position = "none"), 
  McCann_resources    + theme(legend.position = "none"),
  Stability_McCann    + theme(legend.position = "none"), ncol=2, align='hv', labels = "AUTO")

@@

```{r bifur_plot, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:bifur_plot}**Divergent character displacement creates unstable food webs**. I simulated the eco-evolutionary dynamics of character displacement assuming a more realistic functional response for consumers. The results are qualtitatively similar for consumer dynamics (not shown). Note that the lines illustrate the ecological consequences across the range of consumer specialization, while the points are the results of an eco-evolutionary simulation. In this simulation, this symmetry in attack rates may be broken, which is why the point of instability does not exactly match the lines. I stopped the simulations once consumers reached an evolutionary stable strategy, a constraint of maximum specialization, or created an unstable food web. I do not project beyoned this, because the dynamics of mutant invaders becomes unclear when resource distributions are unstable. The grey arrow denotes the boundary of local stability."}
knitr::include_graphics("bifurcation_plot.pdf")
```

@@

>>>>

Previous studies have analyzed the evolution of consumer attack rates in the first two foraging scenarios using an Adaptive Dynamics approach, with the general result being character divergence [@Lawlor1976; @Abrams1986]. 
I say consumers have undergone character divergence if there evolved attack rates are more specialized, measured as $|\frac{a_{ii}}{a_{ii}+a_{ij}}-0.5|$, when evolving with a competing consumer.
I also used an Adaptive Dynamics approach to analyze character displacement in the third foraging scenario, and I too observed character divergence (detailed analysis in Appendix S1 of Supplementary Information).

@@

Given that character divergence was an expected result, I focus here on its consequences for food-web dynamics across these foraging scenarios.
To do this, I analyzed differences in resource abundances and food-web stability at equilibrium. 
An equilibrium is reached when the four rates of change in the above equations are 0, and solving the system at this point gives equilibrium abundances for each resource ($\hat R_1,\hat R_2$) and each consumer ($\hat C_1,\hat C_2$). 
At the coexistence equilibrium (i.e., where $\hat R_1,\hat R_2, \hat C_1,\hat C_2 > 0$), I calculated resource abundances and performed a local stability analysis [@Otto2007]. 
This stability analysis derives the dominant eigenvalue, $\lambda$, which determines whether (and how readily) the food web will return to equilibrium after a small perturbation.

When possible, I derived analytical expressions for the relationship between consumer attack rates and food-web dynamics. 
To do this, I simplified the model by assuming that resources are equivalent ($r=r_1=r_2$ and $K=K_1=K_2$) as well as consumers ($e=e_{11}=e_{12}=e_{21}=e_{22}$; $h=h_{11}=h_{12}=h_{21}=h_{22}$; $m=m_1=m_2$), except that consumer attack rates and their habitat preferences (if present) are mirror images of each other ($a_{ii}=a_{11}=a_{22}$; $a_{ij}=a_{12}=a_{21}$; $w_{ii}=w_{11}=w_{22}$). 
Controlling for other sources of variability allowed me to isolate the general effects of character divergence.
All mathematical derivations were conducted in @Mathematica.

@@

This equation implies that if consumers evolve to become specialists on resources that occur in their preferred habitat (e.g. $w_{ii}>0.5$ and $a_{ii}>a_{ij}$), then the effective attack rate of consumers ($wa_{ii}+(1-w_{ii})a_{ij}$) will always increase, regardless of the tradeoff (Fig. \ref{fig:plot_fig3}). 
Thus, character divergence always results in resource suppression (Fig. \ref{fig:plot_fig3}). 
Note that the shape of the tradeoff can modify the effect of character displacement. 
This is not so much due to the tradeoff affecting the magnitude of displacement (it does, but the effect is minor), but because the form of the tradeoff affects resource abundances when a single consumer has reached an evolutionary stable strategy (ESS, triangles in Fig. \ref{fig:plot_fig3}). 
In contrast, resource abundances reach a similar value when consumers evolve in the presence of a competitor (circles in Fig. \ref{fig:plot_fig3}), because character displacement tends to reach a constraint of complete specialization. 
It is worth noting that resource abundances are consistently higher at the single consumer ESS compared to the predictions I derived for when both consumers are present (deviation of triangles from respective colored lines in Fig. \ref{fig:plot_fig3}). 
This is because consumers actually evolve to be slightly specialized on the resources that occur in their non-preferred habitat (deviation of triangles from 0.5 along x-axis in Fig. \ref{fig:plot_fig3}).

As seen previously, the effect of character displacement on resource abundances qualitatively corresponds to its effect on food-web stability (Fig. \ref{fig:plot_fig3}C). 
Specifically, character divergence decreases food-web stability, regardless of the tradeoff in attack rates.
This is not simply a consequence of having an additional consumer in the system, but emerges from the eco-evolutionary feedback between character displacement and resource suppression (Fig. \ref{fig:plot_fig3}C). 
For example, when the tradeoff is concave up (orange), then the four-species (small circle) food web actually starts off as more stable than the three-species one (small triangle); however, this pattern switches by the end of the eco-evolutionary simulation (large points).

@@

In the third foraging scenario, I observed the same general effect of character divergence as the previous scenario (resources in different habitats, but linear functional response).
<!--Divergent character displacement in this foraging scenario has the same general effect on food-web dynamics as the previous scenario (resources in different habitats, but linear functional response). 
<!--In terms of food-web dynamics, I observe the same general effect of ecological character displacement as before when resources occurred in distinct habitats, but consumers had a simpler functional response. -->
This is because resource abundances at equilibrium are governed by a similar dynamic (derived in Appendix S**?**):
<!--
Even after adding more biological realism, I still observe the same qualitative relationship between character displacement and resource suppression (Fig. \ref{fig:McCann_Plot} in Appendix). This is because resource competition always leads to character divergence and equilibrium resource abundances are governed by similar dynamics  (derivation in supplementary material):
-->

@@

```{r Stability_McCann, include=FALSE, fig.cap="\\label{fig:Stability_McCann}Effect of character displacement on local stability when consumers exhibit a more realistic functional response."}
Stability_McCann <- cbind.data.frame(select(tradeoff_df, n, aii, aij), McCann_2C_2R_tradeoff_dynamics) %>%
  ggplot(., aes(x=aii/(aii+aij), y=-1*max.Re.eigen)) + 
  geom_line(aes(color=factor(n))) +
  #geom_point(data = filter(McCann_displacements, Time=="End"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=3, color="white", show.legend = FALSE) +
  #geom_point(data = filter(McCann_displacements, Time=="Begin"), aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor), size=1.5, color="white") +
  geom_point(data = McCann_displacements, aes(x = a11/(a11+a12), y=-1*max.Re.eigen, fill=factor(n), shape=Competitor, size=Time), color="white", inherit.aes = FALSE) +
  xlab(expression(frac(a[ii],a[ii]+a[ij]))) +
  ylab(bquote(atop("Food-web stability","("*-lambda[max]*")"))) +
  #ylab(expression("Stability "(-lambda[max]))) + 
  scale_shape_manual(values=c(24,21))  +
  scale_size_manual(values=c(1.5,3)) +
  scale_color_manual(name="Tradeoff (n)", values=cbPalette)  +
  scale_fill_manual(name="Tradeoff (n)", values=cbPalette) +
  geom_hline(yintercept=0, linetype="dotted", color="grey") +
  guides(fill = guide_legend(override.aes=list(shape=21)))

@@

```{r McCann Model, include=FALSE}
McCann_legend <- get_legend(Stability_McCann)

plot_McCann_model <- plot_grid(
  McCann_attack    + theme(legend.position = "none") + xlab(""), 
  McCann_resources    + theme(legend.position = "none") + xlab(""),
  Stability_McCann    + theme(legend.position = "none") + xlab(""), ncol=2, align='hv') # , labels = "AUTO"

plot_McCann_model <- add_sub(plot_McCann_model, label = expression(paste("Specialization ", bgroup("(",frac(a[ii],a[ii]+a[ij]),")"))))

#plot_McCann_model <- ggdraw(plot_McCann_model) + draw_grob(McCann_legend, x = 0.7, y=-0.2)

@@

```{r bifur_plot, echo=FALSE, fig.pos="H", fig.cap="\\label{fig:bifur_plot}**Character divergence creates unstable food webs**. I simulated the eco-evolutionary dynamics of character displacement assuming a more realistic functional response for consumers. The results are qualtitatively similar for consumer dynamics (not shown). Note that the lines illustrate the ecological consequences across the range of consumer specialization, while the points are the results of an eco-evolutionary simulation. In this simulation, this symmetry in attack rates may be broken, which is why the point of instability does not exactly match the lines. I stopped the simulations once consumers reached an evolutionary stable strategy, a constraint of maximum specialization, or created an unstable food web. I do not project beyoned this, because the dynamics of mutant invaders becomes unclear when resource distributions are unstable. The grey arrow denotes the boundary of local stability."}
knitr::include_graphics("bifurcation_plot.pdf")
```

@@

```{r all asymmetry}

all_asymm_plots <- plot_grid(
  Mac_attack_asymm + ggtitle("Scenario I\n") + theme(legend.position = "none", axis.text.x=element_blank()) + xlab("") + ylab(bquote(atop("Effective attack rate","("*a[ii]+a[ij]*")"))), #, plot.title=element_text(size=8)
  LS_attack_asymm  + ggtitle("Scenario II\n") + theme(legend.position = "none", axis.text.x=element_blank()) + xlab("") + ylab(bquote(w[ii]*a[ii]+"("*1-w[ii]*")"*a[ij]*")")), #, plot.title=element_text(size=8)
  McCann_attack_asymm + ggtitle("Scenario III")  + theme(legend.position = "none", axis.text.x=element_blank()) + xlab("") + ylab(bquote(w[ii]*a[ii]+"("*1-w[ii]*")"*a[ij]*")")), #, plot.title=element_text(size=8), plot.subtitle=element_text(size=8)
  Mac_resources_asymm   + theme(legend.position = "none", axis.text.x=element_blank()) + xlab("") + ylab(bquote(atop("Resource abundance","("*R[1]+R[2]*")"))),
  LS_resources_asymm   + theme(legend.position = "none", axis.text.x=element_blank()) + xlab("") + ylab(""),
  McCann_resources_asymm   + theme(legend.position = "none", axis.text.x=element_blank()) + xlab("") + ylab(""),
  Stability_Mac_asymm   + theme(legend.position = "none") + scale_x_continuous(labels=c(0,0.25,0.5,0.75,1)) + xlab("") + ylab(bquote(atop("Food-web stability","("*-lambda[max]*")"))),
  Stability_LS_asymm   + theme(legend.position = "none") + scale_x_continuous(labels=c(0,0.25,0.5,0.75,1)) + xlab("") + ylab(""),
  Stability_McCann_asymm   + theme(legend.position = "none") + scale_x_continuous(labels=c(0,0.25,0.5,0.75,1)) + xlab("") + ylab(""),
  ncol=3, align='v'#, rel_widths = c(1,0.9,0.9)
)

all_asymm_plots <- add_sub(all_asymm_plots, label = expression(paste("Specialization ", bgroup("(",frac(a[ii],a[ii]+a[ij]),")"))))

save_plot(filename="asymm_figs.pdf", plot=all_asymm_plots, base_height = 11, base_width = 8.5)
```
